Basic concepts and tools for the Toki Pona minimalist and constructed language: Wordnet synsets; analysis of the vocabulary; synthesis and syntax highlighting of texts

### 2017-12-16 article/article.tex
detail about 24-bit true color

\begin{document}
\title{Basic concepts and tools for the Toki Pona minimalist language:
Wordnet synsets; analysis, synthesis and syntax highlighting of texts}
% Anthropological computation
\author{Renato Fabbri\\
\texttt{renato.fabbri@gmail.com}\\
University of São Paulo,\\
Institute of Mathematical and Computer Sciences\\
São Carlos, SP, Brazil
}
\maketitle
\begin{abstract}
  A minimalist constructed language (conlang)
  is useful for experiments and comfortable for making tools.
  The Toki Pona (TP) conlang is minimalist both in the vocabulary
  (with only 14 letters and 124 words)
  and in the $\approx10$ syntax rules.
  The language is useful for being a used and somewhat established
  minimalist conlang with at least hundreds of fluent speakers.
  In this article, we describe current concepts and resources
  for TP,
  and make available Python scripted routines for
  the analysis of the language,
  the synthesis of texts, the specification of syntax highlighting
  schemes, and the achievement of a preliminary TP
  Wordnet~\cite{wordnet}.
  We focus on the analysis of the basic vocabulary,
  as corpus analyses were found in~\cite{corpus}.
  The synthesis is based on sentence templates,
  relates to context by keeping track of used words,
  and renders larger texts by using a fixed number of phonemes (e.g. for poems)
  and number of sentences, words and letters (e.g. for paragraphs).
  Syntax highlighting 
  reflects morphosyntactic classes given in the official dictionary
  and different solutions are described and implemented
  using 24bit true colors in the well-established Vim text editor~\cite{vim}.
  The tentative TP Wordnet is made available in three forms that
  reflect the choices of the synsets related to each word.
  In summary, this text holds potentially novel conceptualizations about,
  and tools and results in analyzing, synthesizing or syntax highlighting the
  TP language.
  % subj (li), obj (e), prepositions, qualification (word concatenation), 
  % sent (la), particles, mi-sina, pi, punctuation is freestyle, pre-verb,
\end{abstract}
{\bf keywords:} Constructed languages, Natural Language Processing, Syntax highlighting, Wordnet, Toki Pona

@@

All the words except the particles (li, e, la, pi, a, o, anu, en, seme, mu)
and the words classified solely as
prepositions (kepeken, lon, tan)
are usable in any position in noun and verb phrases,
(i.e. 107 words discarding synonyms).
Notice that the phrase expresses a noun (in a subject or
object) or a verb (in a predicate).
And that the first word of the phrase is the noun or verb,
and that subsequent words are adjectives or adverbs.
The pi precedes a noun or a verb to be qualified and then qualify
the phrase that comes before it.

At this point, the only missing syntax rule is related to
the prepositions: kepeken, lon, sama, tan, tawa.
They might appear at the end of a phrase,
should be followed by another phrase,
and require no particle. E.g.
`toki *tan* jan Pije li pana e sona *tawa* mi'.
Because prepositions can be used as adjectives
or nouns or verbs or adverbs, etc, and might follow
anu noun or verb phrase,
they are one of the three main sources of ambiguity:
prepositions; absence of li after sina and mi; incomplete sentences.
E.g. '(mi [li]) pana tawa kon e ilo pi suli mute'.

@@

\begin{itemize}
  \item a or kin, emphasis.
  \item o, vocative or imperative ('jan lukin sitelen o, li wawa')
  \item taso, means however as sentence or 'only' if adjective.
  \item anu, en: 'or' and 'and'. Used for nouns in noun phrases.
    For verbs, repeat li.
    For object nouns, repeat e.
    If the noun is complementing a preposition (e.g. tawa, lon),
    one might repeat the preposition.
    As TP is a recent language, and is particularly able to cope with
    variation due to its simplicity, I would advocate for
    using en and anu wherever there is no ambiguity.
    E.g.: 'mi tawa mute anu moku lon tenpo lili'
    In the official documentation that is not described and
    might be regarded as wrong in strict canonical TP.
  \item nanpa, denotes numbering.
  \item seme, for questions, used next to the thing being asked for.
    'tan seme la sina pana e sike?'.
  \item mu, for animal noises. For me it is not a particle, as in the
    official dictionary, but a noun.
    I also like to use it as a verb:
    'mi pakala e luka. mi mu mute.'
\end{itemize}

@@

\subsubsection{Recognizing POS tags by speaker and machine}
One might perform a POS tagging by following these rules:
\begin{itemize}
  \item Noun: the first word in a noun phrase.
    Starting a sentence if it is complete,
    after an 'e', after a pi, and after a preposition.
    The predicate is often a noun
    if the sentence has no object.
  \item Adjective: second word on after an e and after
    second word on in the subject phrase if present.
  \item Verb: after a li, mi or sina.
    If there is no object, the predicate
    might be a noun or a verb.
  \item After a preposition,
    there is a noun phrase.
  \item Notice that there is ambiguity in the structure
    introduced by the omission of li after mi and sina.
    Also, when there is no object, a noun or a verb
    or an adjective might be in the verb position.
    % The prepositional complement is also not defined.
    % (mi moku tawa pali, tawa tomo). % TTM
    These are sources of syntactic ambiguities
    in TP.
    They might be resolved or minimized by using the semantics
    of the words.
    An initial effort in this direction might be
    using the classes in the official dictionary to resolve
    ambiguities whenever possible.
    This solution is not optimal for correct POS tagging,
    and does not solve all possible ambiguities
    (there are words classified as noun and adjective,
    adjective and verb, noun and verb, particle and verb).
    Another source of ambiguity is the pre-verbs as described
    in the literature~\cite{tpLang,kama}.
    But I find it reasonable to understand them as verbs,
    as described in the end of Section~\ref{sec:par}.
  \item If there is no object, the predicate might be parsed
as being a hybrid of noun, verb and adjective.
  \item The prepositional phrase might also be taken to be a hybrid
    of noun, verb and adjective.
\end{itemize}

@@

\subsubsection{Additional notes}
John Clifford (a notable ``toki-ponist'', a.k.a. jan Kipo)
states that there are not noun, verb or prepositional
phrases in TP, but only phrases in a structure
yield by particles and content words~\cite{janKipo}.
The only synonyms on Toki Pona are:
a and kin; lukin and oko; sin or kamako;
ale or ali.
In formations such as
toki e ni:, wile e ni:, tan ni: etc.,
'(e) ni' might be omitted and : used alone.
Names are by default transliterated,
but might not be.
These and other issues are further described in Appendix~\ref{mytoki}.

@@

Possible sentence structures are unlimited.
For a noun or verb phrase, one might quantify the possibilities
by considering all words but the particles that are not classified
also as something else
(i.e. li, e, la, pi, a, o, anu, en, seme, mu)
and the prepositions that are not classified also as something
else (i.e. kepeken, lon, tan).
This yields $120-13=107 words$.
As discussed in Section~\ref{syntax},
I advocate that, although such words are classified in the dictionary,
they might be used indistinctly as nouns, adjectives,
verbs and adverbs.
Thus, one has 107 possibilities of noun and verb phrases with
one word, $107^2=11,449$ possibilities with two words,
$107^3=1,225,043$ with three words and so on.
Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
(verb), object (noun) and prepositional (noun or verb) phrases of a sentence,
and assume that the sentence has at most one prepositional phrase,
one might quantify the possibilities using the formula:
$\delta = 107^n\times 107^v\times 10^o \times 5\times 107^p$,
where $5$ stands for the possible prepositions.
To account for the particles, one possibility is
to assume for simplicity that one might use none or one particle
(not li, e, la, pi, resulting in $8$ particles and $9$ possibilities)
at each phrase, yielding:
$\delta = 107^n\times 107^v\times 107^o \times 5\times 107^p\times 9^4$.
For example, assume $n=v=o=p=1$, then
$\delta=107^1\times 107^1\times 107^1 \times 5\times 107^1\times 9^4=
 4.300066\times 10^+12$,
i.e. more then 4 trillion possible
sentences with single word phrases (i.e. no adjectives or adverbs),
while allowing only one particle per phrase
and only one predicate phrase
(e.g. 'ona li moku e soweli lon supa').
The general case is implemented as a function
in the script \tttt{makeStatistics.py}~\cite{tokipona}.

@@

I have been inclined to conceive a language with the same syntax of TP:
<sentence> la <sentence = subject + predicate + object (each term with
a possible preposition complement)>
but always using particles to bound the sentence sections,
repeating them when the function is performed in the smallest scale:
'toki pi toki pona pi jan sona' meaning
language in toki pona and of intellectuals vs
'toki pi toki pona pi pi jan sona'
meaning language in [toki pona of intelectuals].
About the keyword for initiating a preposition complement 
might be 'a', but it is already taken in TP.
Maybe use 'a' before a preposition and use 'ha' or 'he'
instead of TP a.
In such a setting,
one might enable pi li e a (concept-qualifier, subject-predicate,
predicate-object, concept-preposition)
in any slot of the template:
<sentence> la <sentence = subject + predicate + object (each term with
a possible preposition complement)>
Between sentences one would use:
la pi, la li, la e, la a, and la la.
This is a very fractal conlang proposal.
Maybe also have a way to discern between a noun,
a qualifier (adjectives/adverbs) and a verb,
and accept any of them for the subject, predicate, object, preposition,
slots. Or assume a part of speech as default for a slot or for the
first word in a slot.

@@

o taso la toki pona li kalama li lukin pona.
sitelen en nasin li open e sitelen
suli\cite{tpLang,kama,akesiWawa,gdoc,tokisona,Wikipesija}.
li sona. li nasin e toki e sona e lawa e lon.

@@

  Clifford, J. (2017, December 15). Considerations about Toki Pona

>>>>

\begin{document}
\title{Basic concepts and tools for the Toki Pona minimalist and
constructed language:
Wordnet synsets; analysis of the vocabulary; synthesis and syntax highlighting of texts}
% Anthropological computation
\author{Renato Fabbri\\
\texttt{renato.fabbri@gmail.com}\\
University of São Paulo,\\
Institute of Mathematical and Computer Sciences\\
São Carlos, SP, Brazil
}
\maketitle
\begin{abstract}
A minimalist constructed language (conlang)
is useful for experiments and comfortable for making tools.
The Toki Pona (TP) conlang is minimalist both in the vocabulary
(with only 14 letters and 124 words)
and in the $\approx10$ syntax rules.
The language is useful for being a used and somewhat established
minimalist conlang with at least hundreds of fluent speakers.
In this article, we describe current concepts and resources
for TP,
and make available Python scripted routines for
the analysis of the language,
the synthesis of texts, the specification of syntax highlighting
schemes, and the achievement of a preliminary TP
Wordnet~\cite{wordnet}.
We focus on the analysis of the basic vocabulary,
as corpus analyses were found in~\cite{corpus}.
The synthesis is based on sentence templates,
relates to context by keeping track of used words,
and renders larger texts by using a fixed number of phonemes (e.g. for poems)
and number of sentences, words and letters (e.g. for paragraphs).
Syntax highlighting 
reflects morphosyntactic classes given in the official dictionary
and different solutions are described and implemented
in the well-established Vim text editor~\cite{vim}.
The tentative TP Wordnet is made available in three forms that
reflect the choices of the synsets related to each word.
In summary, this text holds potentially novel conceptualizations about,
and tools and results in analyzing, synthesizing or syntax highlighting the
TP language.
  % subj (li), obj (e), prepositions, qualification (word concatenation), 
  % sent (la), particles, mi-sina, pi, punctuation is freestyle, pre-verb,
\end{abstract}
{\bf keywords:} Constructed languages, Natural Language Processing, Syntax highlighting, Wordnet, Toki Pona

@@

All the words except the particles (li, e, la, pi, a, o, anu, en, seme, mu)
and the words classified solely as
prepositions (kepeken, lon, tan)
are usable after any noun or verb phrase,
(i.e. 107 words discarding synonyms).
Notice that the phrase expresses a noun (in a subject or
object) or a verb (in a predicate).
And that the first word of the phrase is the noun or verb,
and that subsequent words are adjectives or adverbs.
The pi precedes a noun or a verb to be qualified and then qualify
the phrase that comes before it.

At this point, the only missing syntax rule is related to
the prepositions: kepeken, lon, sama, tan, tawa.
They might appear at the end of a phrase,
should be followed by another phrase,
and require no particle. E.g.
`toki *tan* jan Pije li pana e sona *tawa* mi'.
Because prepositions can be used as adjectives
or nouns or verbs or adverbs, etc, and might follow
any noun or verb phrase,
they are one of the three main sources of ambiguity:
prepositions; absence of li after sina and mi; incomplete sentences.
E.g. '(mi [li]) pana tawa kon e ilo pi suli mute'.

@@

\begin{itemize}
  \item a or kin, emphasis.
  \item o, vocative or imperative ('jan lukin sitelen o, li wawa')
  \item taso, means however as sentence or 'only' if adjective.
  \item anu, en: 'or' and 'and'. Used for nouns in noun phrases.
    For verbs, repeat li.
    For object nouns, repeat e.
    If the noun is complementing a preposition (e.g. tawa, lon),
    one might repeat the preposition.
    As TP is a recent language, and is particularly able to cope with
    variation due to its simplicity, I would advocate for
    using en and anu/en wherever there is no ambiguity.
    E.g.: 'mi tawa mute anu moku lon tenpo lili'
    In the official documentation that is not described and
    might be regarded as wrong in strict canonical TP.
  \item nanpa, denotes numbering.
  \item seme, for questions, used next to the thing being asked for.
    'tan seme la sina pana e sike?'.
  \item mu, for animal noises. For me it is not a particle, as in the
    official dictionary, but a noun.
    I also like to use it as a verb:
    'mi pakala e luka. mi mu mute.'
\end{itemize}

@@

\subsubsection{Recognizing POS tags by speaker and machine}\label{sec:rec}
One might perform a POS tagging by following these rules:
\begin{itemize}
  \item Noun: the first word in a noun phrase.
    Starting a sentence if it is complete,
    after an 'e', after a pi.
    The predicate is often a noun
    if the sentence has no object.
  \item Adjective: second word on after an e and after
    second word on in the subject phrase if present.
  \item Verb: after a li, mi or sina.
    If there is no object, the predicate
    might be a noun or a verb.
  \item After a preposition, one might consider that what follows is 1) always
    a noun, or that 2) it can be a noun, adjective or verb,
    or that 3) it is a hybrid of noun, verb and adjective.~\cite{janKipo}
  \item Notice that there is ambiguity in the structure
    introduced by the omission of li after mi and sina.
    Also, when there is no object, a noun or a verb
    or an adjective might be in the verb position.
    % The prepositional complement is also not defined.
    % (mi moku tawa pali, tawa tomo). % TTM
    These are sources of syntactic ambiguities
    in TP.
    They might be resolved or minimized by using the semantics
    of the words.
    An initial effort in this direction might be
    using the classes in the official dictionary to resolve
    ambiguities whenever possible.
    This solution is not optimal for correct POS tagging,
    and does not solve all possible ambiguities
    (there are words classified as noun and adjective,
    adjective and verb, noun and verb, particle and verb).
    Another source of ambiguity is the pre-verbs as described
    in the literature~\cite{tpLang,kama}.
    But I find it reasonable to understand them as verbs,
    as described in the end of Section~\ref{sec:par}.
  \item If there is no object, the predicate might be parsed
as being a hybrid of noun, verb and adjective.
\end{itemize}

@@

\subsubsection{Additional notes}\label{sec:add}
John Clifford (a notable ``toki-ponist'', a.k.a. jan Kipo)
states that there are not noun, verb, adjective (or even prepositional)
phrases in TP, but only phrases in a structure
yield by particles and content words~\cite{janKipo}.
The only synonyms on Toki Pona are:
a and kin; lukin and oko; sin or namako;
ale or ali.
In formations such as
toki e ni:, wile e ni:, tan ni: etc.,
'(e) ni' might be omitted and : used alone.
Names are by default transliterated,
but might not be.
These and other issues are further described in Appendix~\ref{mytoki}.

@@

Possible sentence structures are unlimited.
For a noun or verb phrase, one might quantify the possibilities
by considering all words but the particles that are not classified
also as something else
(i.e. li, e, la, pi, a, o, anu, en, seme, mu)
and the prepositions that are not classified also as something
else (i.e. kepeken, lon, tan).
This yields $120-13=107 words$.
As discussed in Section~\ref{syntax},
I advocate that, although such words are classified in the dictionary,
they might be used indistinctly as nouns, adjectives,
verbs and adverbs.
Thus, one has 107 possibilities of noun and verb phrases with
one word, $107^2=11,449$ possibilities with two words,
$107^3=1,225,043$ with three words and so on.
Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
(verb), object (noun) and prepositional (see Sections~\ref{sec:rec} and~\ref{sec:add}) phrases of a sentence,
and assume that the sentence has at most one prepositional phrase,
one might quantify the possibilities using the formula:
$\delta = 107^n\times 107^v\times 10^o \times 5\times 107^p$,
where $5$ stands for the possible prepositions.
To account for the particles, one possibility is
to assume for simplicity that one might use none or one particle
(not li, e, la, pi, resulting in $8$ particles and $9$ possibilities)
at each phrase, yielding:
$\delta = 107^n\times 107^v\times 107^o \times 5\times 107^p\times 9^4$.
For example, assume $n=v=o=p=1$, then
$\delta=107^1\times 107^1\times 107^1 \times 5\times 107^1\times 9^4=
 4.300066\times 10^+12$,
i.e. more then 4 trillion possible
sentences with single word phrases (i.e. no adjectives or adverbs),
while allowing only one particle per phrase
and only one predicate phrase
(e.g. 'ona li moku e soweli lon supa').
The general case is implemented as a function
in the script \tttt{makeStatistics.py}~\cite{tokipona}.

@@

I have been inclined to conceive a language with the same syntax of TP:
'sentence' la 'sentence = subject + predicate + object (each term with
a possible preposition complement).
but always using particles to bound the sentence sections,
repeating them when the function is performed in the smallest scale:
'toki pi toki pona pi jan sona' meaning
language in toki pona and of intellectuals vs
'toki pi toki pona pi pi jan sona'
meaning language in [toki pona of intelectuals].
About the keyword for initiating a preposition complement 
might be 'a', but it is already taken in TP.
Maybe use 'a' before a preposition and use 'ha' or 'he'
instead of TP a.
In such a setting,
one might enable pi li e a (concept-qualifier, subject-predicate,
predicate-object, concept-preposition)
in any slot of the template:
'sentence' la 'sentence = subject + predicate + object (each term with
a possible preposition complement)'
Between sentences one would use:
la pi, la li, la e, la a, and la la.
This is a very fractal conlang proposal.
Maybe also have a way to discern between a noun,
a qualifier (adjectives/adverbs) and a verb,
and accept any of them for the subject, predicate, object, preposition,
slots. Or assume a part of speech as default for a slot or for the
first word in a slot.

@@

o taso la toki pona li kalama li lukin pona.
sitelen en nasin li open e sitelen
suli~\cite{tpLang,kama,akesiWawa,gdoc,tokisona,Wikipesija}.
li sona. li nasin e toki e sona e lawa e lon.

@@

  Clifford, J. et al. (2017, December 15). Considerations about Toki Pona

### 2017-12-16 article/pos.tex
detail about 24-bit true color

\begin{table*}[h!]
\begin{center}
\caption{POS tags incident and chosen.
            The official dictionary often relates tokens
            to more than one POS tag.
            For the current version of the text highlighting plugin described in
            Section~\ref{shigh}, for example,
            a token has to have an established tag to have
            a defined color.
            On the 'Chosen' column, the tokens were regarded only once
            by choosing the class in the dictionary respecting the
            precedence order: PRE, VERB, PREPOSITION, PARTICLE, ADJECTIVE, NOUN, NUMBER.}\label{foobar}
\begin{tabular}{| l | c | c |}\hline
POS & All  & Chosen \\\hline
NOUN & 58  & 49 \\\hline
ADJECTIVE & 40  & 34 \\\hline
VERB & 15  & 13 \\\hline
PARTICLE & 12  & 12 \\\hline
PRE & 6  & 6 \\\hline
PREPOSITION & 5  & 5 \\\hline
NUMBER & 4  & 1 \\\hline
total & 140  & 120 \\\hline
\end{tabular}\end{center}
\end{table*}

>>>>

\begin{table*}[h!]
\begin{center}
\caption{POS tags incident and chosen.
            The official dictionary often relates tokens
            to more than one POS tag.
            For the current version of the text highlighting plugin described in
            Section~\ref{shigh}, for example,
            a token has to have an established tag to have
            a defined color.
            On the 'Chosen' column, the tokens were regarded only once
            by choosing the class in the dictionary respecting the
            precedence order: PRE-VERB, VERB, PREPOSITION, PARTICLE, ADJECTIVE, NOUN, NUMBER.}\label{foobar}
\begin{tabular}{ l | c c }
POS & All  & Chosen \\\hline
NOUN & 58  & 49 \\
ADJECTIVE & 40  & 34 \\
VERB & 15  & 13 \\
PARTICLE & 12  & 12 \\
PRE-VERB & 6  & 6 \\
PREPOSITION & 5  & 5 \\
NUMBER & 4  & 1 \\\hline
total & 140  & 120 \\
\end{tabular}\end{center}
\end{table*}
